COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-24 TACO-paper/coffee.tex
Minor adjustments

\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfig}
\usepackage{tabulary}

@@

\category{G.1.8}{Numerical Analysis}{Partial Differential Equations - Finite element methods}

@@

\begin{algorithm}[t]
\SetAlgorithmName{LISTING}{}
\footnotesize
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do generalized loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// $load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[0,0], A[1,1], A[2,2], A[3,3] \\
~~~~~~~~// One $permute\_pd$ intrinsic per \texttt{k}-loop $load$\\
~~~~~~~~// Compute A[0,1], A[1,0], A[2,3], A[3,2] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsic per \texttt{k}-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz problem after application of \emph{op-vect} on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}. Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length. The original size of the \texttt{j}-\texttt{k} iteration space (i.e. before padding was applied) was 6$\times$6. In this example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\end{algorithm}

@@

\begin{algorithm}
\SetAlgorithmName{LISTING}{}
\footnotesize
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~~~// Same code as in Listing~\ref{code:helmholtz-licm} up to line 15 \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz problem in which \emph{split} has been applied on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
\label{code:helmholtz-split}
\end{algorithm}

@@

\begin{figure}[t]
\begin{alltt}
\footnotesize
\textbf{Input:} ast: abstract syntax tree of a local assembly kernel produced by FFC
       wrapper: the function within which the kernel is invoked
	       isa: represents the SIMD Instruction Set Architecture
\textbf{Output:} assembly code (C string)
\textbf{// Analyze ast and build optimization plan }
it\_space = analyze(ast) 
\textbf{if not} it\_space
  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
  \textbf{return} wrapper + ast.from\_ast\_to\_c()
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) 
\textbf{// Optimize ast based on plan}
ast.licm()
ast.padding() 
ast.align\_data()
\textbf{if} plan.permute
  ast.permute\_assembly\_loops()
\textbf{if} plan.split\_factor
  ast.split(plan.split\_factor)
\textbf{if} plan.uaj\_factor
  uaj = MIN(plan.uaj\_factor, it\_space.j.size/isa.vf) 
  ast.op\_vect(uaj)
\textbf{return} wrapper + ast.from\_ast\_to\_c()
\end{alltt}

\caption{Outline of the COFFEE architecture.}
\label{algo:PyOP2Compiler}
\end{figure}

@@

% "vendor compilers such as Intel's" is awkward and unclear. If Intel is the only one you have looked at, then say "the Intel compiler", otherwise name the ones for which you know this to be true.
Since loop unroll and unroll-and-jam of outer loops are fundamental to the exposure of instruction-level parallelism, tuning critical parameters such as the unroll factor is of great importance. However, inspecting assembly code and comparing with other hand-written implementations indicates that recent versions of the Intel compiler are capable of estimating close-to-optimal values for such parameters. This is particularly true for assembly kernels, in which the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are stride one. We therefore leave the backend compiler in charge of selecting the unroll factor. This choice also simplifies the COFFEE's cost model. The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added.

@@

\begin{figure}
\begin{alltt}
\footnotesize
\textbf{Input:} n\_outer\_arrays, n\_inner\_arrays, n\_consts, n\_regs
\textbf{Output:} uaj\_factor, split\_factor
n\_outer\_regs = n\_regs / 2 
split\_factor = 0 
\textbf{// Compute spltting factor }
\textbf{while} n\_outer\_arrays > n\_outer\_regs
  n\_outer\_arrays = n\_outer\_arrays / 2 
  split\_factor = split\_factor + 1
\textbf{// Compute unroll-and-jam factor for \emph{op-vect}}
n\_regs\_avail = n\_regs - (n\_outer\_arrays + n\_consts)
uaj\_factor = n\_reg\_avail / n\_inner\_arrays
\textbf{// Estimate the benefit of permuting loops}
permute = n\_outer\_arrays > n\_inner\_arrays
\textbf{return} <permute, split\_factor, uaj\_factor>
\end{alltt}
\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when \emph{op-vect} is used) and split factors, avoiding the overhead of auto-tuning.}
\label{algo:applyCostModel}
\end{figure}

The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes into account the number of available logical vector registers, \emph{n\_regs}, and the number of unique variables accessed: \emph{n\_consts} counts variables independent of both \texttt{j} and \texttt{k} loops and temporary registers, \emph{n\_outer\_arrays} counts \texttt{j}-dependent variables, and \emph{n\_inner\_arrays} counts \texttt{k}-dependent variables, assuming the \texttt{ijk} loop order. These values are used to estimate unroll-and-jam and split factors for \emph{op-vect} and \emph{split}. If a factor is 0, then the corresponding transformation is not applied. The \emph{split} transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2). A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions. Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size. The unroll-and-jam factor for the \emph{op-vect} transformation is determined as a function of the available logical registers, i.e. those not used for storing hoisted terms (line 9-11). Finally, the profitability of loop interchange is evaluated (line 12-16).

@@

\subsection{The Helmholtz, Diffusion, and Burgers equations}
\label{sec:real-examples}
Our code transformations were evaluated in three real-world problems based on the following PDEs
\begin{enumerate}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{enumerate}
The three chosen benchmarks are completely real examples and comprise the core operators in some of the most frequently encountered finite element problems in scientific computing. The Helmholtz and Diffusion kernels are archetypal second order elliptic operators. They are complete and unsimplified examples of the operators which occur for diffusion and viscosity in fluids, and for imposing pressure in compressible fluids. For instance, they are both extensively used in climate modeling. Very similar operators, for which the same optimisations are expected to be equally effective, apply to elasticity problems, which are at the core of computational structural mechanics. The Burgers kernel is a typical example of a first order hyperbolic conservation law, which occurs in real applications whenever a quantity is transported by a fluid (the momentum itself, in our case). We chose this particular kernel since it applies to a vector-valued quantity, while the elliptic operators apply to scalar quantities. In this way, the operators we have selected as examples are characteristic of both the second and first order operators which dominate in fluids and solids simulations.

The code was written in UFL (available at~\cite{ufl-code}) and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Listing~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel code, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a non-linear problem employing differential operators different from those of Helmholtz and relying on vector-valued quantities, which has a major impact on the generated assembly code (see Listing~\ref{code:burgers}), where a larger number of basis function arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are generated.

>>>>

\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfig}
\usepackage{tabulary}
\usepackage{lineno}

@@

\category{G.1.8}{Numerical Analysis}{Partial Differential Equations - Finite element methods}
\category{G.4}{Mathematical Software}{Parallel and vector implementations}

@@

\begin{algorithm}[t]
\SetAlgorithmName{LISTING}{}
\footnotesize
\KwSty{void} helmholtz(double A[4][4], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~// Do generalized loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k+=4) $\lbrace$\\
~~~~~~~~// $load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[0,0], A[1,1], A[2,2], A[3,3] \\
~~~~~~~~// One $permute\_pd$ intrinsic per \texttt{k}-loop $load$\\
~~~~~~~~// Compute A[0,1], A[1,0], A[2,3], A[3,2] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsic per \texttt{k}-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz problem after application of \emph{op-vect} on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}. In this example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\end{algorithm}

@@

\begin{algorithm}
\SetAlgorithmName{LISTING}{}
\footnotesize
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~~~// Same code as in Listing~\ref{code:helmholtz-licm} up to the j loop\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz problem in which \emph{split} has been applied on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
\label{code:helmholtz-split}
\end{algorithm}

@@

\begin{figure}[t]
\begin{alltt}
\footnotesize
\textbf{Input:} ast: abstract syntax tree of a local assembly kernel produced by FFC
       wrapper: the function within which the kernel is invoked
	       isa: represents the SIMD Instruction Set Architecture
\textbf{Output:} assembly code (C string)
\textbf{// Analyze ast and build optimization plan }
it\_space = analyze(ast) 
\textbf{if not} it\_space
  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
  \textbf{return} wrapper + ast.from\_ast\_to\_c()
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) 
\textbf{// Optimize ast based on plan}
ast.licm()
ast.padding() 
ast.align\_data()
\textbf{if} plan.permute
  ast.permute\_assembly\_loops()
\textbf{if} plan.split\_factor
  ast.split(plan.split\_factor)
\textbf{if} plan.uaj\_factor
  uaj = MIN(plan.uaj\_factor, it\_space.j.size/isa.vf) 
  ast.op\_vect(uaj)
\textbf{return} wrapper + ast.from\_ast\_to\_c()
\end{alltt}
\caption{Outline of the COFFEE architecture.}
\label{algo:PyOP2Compiler}
\end{figure}

@@

% "vendor compilers such as Intel's" is awkward and unclear. If Intel is the only one you have looked at, then say "the Intel compiler", otherwise name the ones for which you know this to be true.
Loop unroll  (or unroll-and-jam of outer loops) is fundamental to the exposure of instruction-level parallelism, and tuning the unroll factor is particularly important. However, we noticed, by comparison with implementations having manually-unrolled loops, that recent versions of the Intel compiler estimate close-to-optimal unroll factors when the loops are affine and their bounds are small and known at compile-time, which is the case of our kernels. We therefore leave the backend compiler in charge of selecting unroll factors. This choice also simplifies the COFFEE's cost model. The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code prevents the Intel compiler from applying this optimization, even if specific pragmas are added.

@@

\begin{figure}
\begin{alltt}\footnotesize\internallinenumbers
\textbf{Input:} n\_outer\_arrays, n\_inner\_arrays, n\_consts, n\_regs
\textbf{Output:} uaj\_factor, split\_factor
n\_outer\_regs = n\_regs / 2 
split\_factor = 0 
\textbf{// Compute spltting factor }
\textbf{while} n\_outer\_arrays > n\_outer\_regs
  n\_outer\_arrays = n\_outer\_arrays / 2 
  split\_factor = split\_factor + 1
\textbf{// Compute unroll-and-jam factor for \emph{op-vect}}
n\_regs\_avail = n\_regs - (n\_outer\_arrays + n\_consts)
uaj\_factor = n\_reg\_avail / n\_inner\_arrays
\textbf{// Estimate the benefit of permuting loops}
permute = n\_outer\_arrays > n\_inner\_arrays
\textbf{return} <permute, split\_factor, uaj\_factor>
\end{alltt}
\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when \emph{op-vect} is used) and split factors, avoiding the overhead of auto-tuning.}
\label{algo:applyCostModel}
\end{figure}

The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes into account the number of available logical vector registers, \emph{n\_regs}, and the number of unique variables accessed: \emph{n\_consts} counts variables independent of both \texttt{j} and \texttt{k} loops and temporary registers, \emph{n\_outer\_arrays} counts \texttt{j}-dependent variables, and \emph{n\_inner\_arrays} counts \texttt{k}-dependent variables, assuming the \texttt{ijk} loop order. These values are used to estimate unroll-and-jam and split factors for \emph{op-vect} and \emph{split}. If a factor is 0, then the corresponding transformation is not applied. The \emph{split} transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 3). A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions. Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size. The unroll-and-jam factor for the \emph{op-vect} transformation is determined as a function of the available logical registers, i.e. those not used for storing hoisted terms (line 9-11). Finally, the profitability of loop interchange is evaluated (line 13).

@@

\subsection{The Helmholtz, Diffusion, and Burgers equations}
\label{sec:real-examples}
Our code transformations were evaluated in three real-world problems based on the following PDEs
\begin{enumerate}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{enumerate}
The three chosen benchmarks are completely real examples and comprise the core differential operators in some of the most frequently encountered finite element problems in scientific computing. The Helmholtz and Diffusion kernels are archetypal second order elliptic operators. They are complete and unsimplified examples of the operators occurring for diffusion and viscosity in fluids, and for imposing pressure in compressible fluids. As such, they are both extensively used in climate and ocean modeling. Very similar operators, for which the same optimisations are expected to be equally effective, apply to elasticity problems, which are at the base of computational structural mechanics. The Burgers kernel is a typical example of a first order hyperbolic conservation law, which occurs in real applications whenever a quantity is transported by a fluid (the momentum itself, in our case). We chose this particular kernel since it applies to a vector-valued quantity, while the elliptic operators apply to scalar quantities; this impacts the generated code, as explained next. In this way, the operators we have selected are characteristic of both the second and first order operators that dominate in fluids and solids simulations, and a wide variety on local assembly codes can be tested.

The benchmarks were written in UFL (code available at~\cite{ufl-code}) and executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Listing~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel code, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix. Burgers is a non-linear problem employing differential operators different from those of Helmholtz and relying on vector-valued quantities, which has a major impact on the generated assembly code (see Listing~\ref{code:burgers}), where a larger number of basis function arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are generated.
